---
title: "Heart-Failure-Survival Analysis-Project"
author: "Martial_Mallika_Lakshmi"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 70
---
Survival Analysis model for predicting mortality caused by Heart Failure.

Introduction : Cardiovascular diseases claim the lives of approximately
17 million people worldwide each year, primarily manifesting as heart
attacks and heart failure. Heart failure (HF) occurs when the heart is
unable to pump sufficient blood to meet the body's needs.

Electronic medical records (EMRs) of patients provide a wealth of data,
including symptoms, physical characteristics, and clinical laboratory
test results. This data can be analyzed using biostatistics to uncover
patterns and correlations that might be missed by medical professionals.

Machine learning, in particular, offers powerful tools for predicting
patient survival based on their medical data. It can also identify the
most critical features within these records, providing valuable insights
for healthcare providers.

Dataset : The dataset includes cardiovascular medical records from 299
patients, consisting of 105 women and 194 men aged between 40 and 95
years. All patients were diagnosed with systolic dysfunction of the left
ventricle and had a history of heart failure. Consequently, each patient
was classified into either class III or class IV of the New York Heart
Association (NYHA) classification, indicating various stages of heart
failure.

#Load survival library
```{r}
suppressMessages(library(survival))
suppressMessages(library(tidyverse))
suppressMessages(library(lubridate))
suppressMessages(library(broom))
suppressMessages(library(ggfortify))
suppressMessages(library(survminer))
```
#Load the dataset, EDA : This section aims to explore dataset data before performing any kind of analysis.
```{r}
setwd("C:/Users/selva/Documents/DSTI/Survival Analysis")
data<-read.csv("heart_failure_clinical_records_dataset.csv")
summary(data)
```
we can see that there are some continous variable that should be
categorical like anaemia, diabetes, high_blood_pressure, sex , smoking
and DEATH_EVENT Let's categorize them



# Categorize variables
# Anaemia
```{r}
df= data %>%
  mutate(anaemia = factor(anaemia, levels = c(0, 1), labels = c("N", "Y")),
         diabetes =factor(diabetes, levels = c(0, 1), labels = c("N", "Y")),
         high_blood_pressure = factor(high_blood_pressure, levels = c(0, 1), labels = c("N", "Y")),
         sex =factor(sex, levels = c(0, 1), labels = c("F", "M")),
         smoking=factor(smoking, levels = c(0, 1), labels = c("N", "Y")),
         DEATH_EVENT=factor(DEATH_EVENT, levels = c(0, 1), labels = c("CENSOR", "EVENT")))

par(mfrow =c(2,3))
barplot(table(df$anaemia), main = "Anaemia", col = "lightblue")
barplot(table(df$diabetes), main = "Diabetes", col = "lightgreen")
barplot(table(df$high_blood_pressure), main = "High Blood Pressure", col = "lightcoral")
barplot(table(df$sex), main = "Sex", col = "lightgoldenrod")
barplot(table(df$smoking), main = "Smoking", col = "lightpink")
barplot(table(df$DEATH_EVENT), main = "EVENT", col = "lightgray")
```
```{r}
str(df)
```
# Basic Non-Parametric estimation of survival : Here we are performing basic analysis using **Kaplan-Meier** methods.
```{r}
kmsurvival = survfit(Surv(df$time, df$DEATH_EVENT) ~ 1)
#summary(kmsurvival)
```

```{r}
autoplot(kmsurvival, xlab="days", ylab="Survival Probability", surv.linetype = 'dashed', 
         surv.colour = 'blue',
         conf.int.fill = 'dodgerblue3', conf.int.alpha = 0.5)+
  labs(title = "Kaplan-Meier Survival Curve", subtitle = "With 95% Confidence Interval") 
```
The survival curve starts at 100% and gradually declines, indicating
that the probability of survival decreases over time as more events
(deaths) occurs.

# Non-Parametric Groups Analysis : This section aims to perform different group analysis on the dataset. 
# Gender Analysis It seems from the analysis that males are living longer than females
```{r}
kmsurvival_gender = survfit(Surv(df$time, df$DEATH_EVENT) ~ df$sex)
#summary(kmsurvival_gender)
```

```{r}
autoplot(kmsurvival_gender,
         censor.shape = '*', facets = TRUE, ncol = 2, xlab="Time", ylab="Survival Probability")
```
Initially, both groups have high survival probabilities, but as time
progresses, differences become more apparent. Males generally have a
higher survival probability compared to females at the early stages but
this might change at later stages depending on the number of events and
sample size. The number of events (deaths) varies over time. For
example, at time 10, females had 1 death while males had 5 deaths,
showing that more deaths occurred in males at this specific time point.
```{r}
# Perform the Log-rank test
logrank_test <- survdiff(Surv(time, DEATH_EVENT == "EVENT") ~ sex, data = df)

# Print the results
logrank_test
```
The log-rank test suggests that there is no evidence to reject the null
hypothesis that the survival curves for males and females are the same.
This is consistent with the findings from the Cox proportional hazards
model, where the effect of sex on survival was also not statistically
significant.
```{r}
# Perform the Log-rank test for the DEATH_EVENT effect
logrank_test_death <- survdiff(Surv(time, DEATH_EVENT == "EVENT") ~ DEATH_EVENT, data = df)
# Print the log-rank test results
print(logrank_test_death)
```
This test doesn't provide new insights because it compares the survival
outcomes based on the outcome itself, which is inherently circular. The
extremely small p-value confirms this obvious distinction between the
two groups.

# Semi-parametric Cox regression
```{r}
# Fit the Cox Proportional Hazards model
cox_model <- coxph(Surv(time, DEATH_EVENT == "EVENT") ~ anaemia + diabetes + high_blood_pressure 
                   + sex + smoking, data = df)
# Summarize the Cox model
summary(cox_model)
```
High Blood Pressure (high_blood_pressureY) is the only predictor with a
statistically significant effect on the hazard of the event, with a
hazard ratio of approximately 1.56. This suggests that individuals with
high blood pressure have a 56% higher risk of the event (DEATH_EVENT)
compared to those without, controlling for other factors in the model.

Other predictors (anaemia, diabetes, sex, smoking) did not show
statistically significant associations with the event at the
conventional 0.05 significance level. The model's concordance index
(0.583) indicates modest predictive ability.

```{r}
# Extract the terms from the Cox model
model_terms <- terms(cox_model)
# Get the variable names
variable_names <- attr(model_terms, "term.labels")
print(variable_names)
```
```{r}
# Use ggforest to plot the Cox model coefficients
ggforest(cox_model)
```
```{r}
# to visualize the coefficients with error bars:
library(ggplot2)
# Extract coefficients and standard errors from the Cox model
cox_summary <- summary(cox_model)
coefficients <- cox_summary$coefficients
# Create a data frame with the necessary information
cox_coef_df <- data.frame(
  Covariate = rownames(coefficients),
  coef = coefficients[, "coef"],
  se_coef = coefficients[, "se(coef)"]
)
# Plot the coefficients with error bars
ggplot(cox_coef_df, aes(x = Covariate, y = coef)) + 
  geom_point() + 
  geom_errorbar(aes(ymin = coef - se_coef, ymax = coef + se_coef), width = 0.2) + 
  labs(title = "Cox Proportional Hazards Model", x = "Covariates", y = "Coefficients") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```
Summary:

anaemiaY: The coefficient is positive and the confidence interval does
not include zero, indicating that anemia is likely associated with an
increased hazard (higher risk) of the event occurring.

high_blood_pressureY : The coefficient is positive and the confidence
interval does not include zero, indicating that its likely associated
with an increased hazard (higher risk) of the event occurring.

diabetesY, sexM, smokingY: The confidence intervals for these covariates
all are below zero, suggesting that their effects are not statistically
significant in this model. This means that these covariates may not have
a strong impact on the hazard rate in this dataset.

Hence high_blood_pressureY and anemia appears to be a significant risk
factor in this Cox proportional hazards model, whereas the other
variables do not show a statistically significant effect based on this
plot.

The limitations of the study is a small dataset, which restricts us to
get more insightful information.

Furthermore,  good awareness programs that  communicate  the  benefits 
of  a  heart  healthy diet and physically active lifestyle could prove 
to be effective and may help to reduce the mortality rate due to heart 
failure conditions.
